Investigating Determinism of Physics Engines for usage in CAV and V\&V simulations

### 2021-08-11 rebuttal_letter/rebuttal.tex

\textit{Comment 5: "The numbers of references are odd"}.
Please let us know any format changes that are required specifically. If you are referring to the document hyperlinks that link to each reference, these can be switched off if required.

\textit{Comment 6: "In section I, the sentence “without the need for millions of miles of costly on-road testing” is too arbitrary, the on-road testing is still necessary for AVs verification"}.
We agree with the reviewer that the need for on-road testing is an essential component of the verification process. However, the previous sentence highlights the shortcomings of on-road testing where human fatalities have occured due to a lack of exploration of the parameter space.
The point of this sentence  was in the context of parameter space exploration and the fact that this space could never be explored entirely with on-road testing alone and hence the neccessity of simulation for verification.

\textit{Comment 7: "No need to provide such a detailed description of floating-point arithmetic since it does not cause non-zero simulation variance"}.
Floating point arithmetic does not contribute to simulation variance, but it is a common misconception that people often jumpt to when considering or discussing deterministic computation. We therefore believe it is important to fully explain the issue to ensure that there is no doubt going forward from this point.

To make this clearer we have modified this section to indicate this misconception. We have also made this section more concise. Please see page 4-5 starting \textit{A common misconception that is often jumped to when concerning non-determinism is the use of floating point number representation. This is an erroneous conflation with non-deterministic computational execution and we explain the reasons for this here}.

\textit{Comment 8: "There is no equal sign in Equation (1)."}.
Thank you we have created a symbol for this to turn this into an equation. See the amended equation on page ...

### 2021-08-11 document/source/body.tex
rebuttal letter 2

Floating-point representation of real numbers is limited by the fixed bit width available in the hardware, resulting in the finite precision with which numbers can be represented in a computation. 
Thus, the results of arithmetic operations must fit into the given bit width, which is achieved through rounding to the nearest representable value. This gives rise to rounding errors~\cite{FloatingPointsBook,goldberg1991every}. %

As a consequence, floating-point arithmetic is not associative. Thus, for arithmetic expressions beyond two operands, such as $x+y+z$, executing $x+(y+z)$ rather than $(x+y)+z$, can produce different results~\cite{Kapre2007}. This is because rounding is being performed at different stages in the computation, depending on the respective execution order. 
%
Execution order changes can occur due to the use of a different compiler or as a consequence of parallelisation, e.g.\ within the runtime environment or at the hardware level. 
%
Furthermore, floating-point calculation results may differ if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.

@@

However, as illustrated above on the example of non-associative addition, a change of execution order can lead to sequences of floating-point operations producing different results. Thus, any uncontrolled options to modify execution order, e.g.\ at the compiler level, within the runtime environment or at the hardware level, can cause simulation results to differ between runs, resulting in non-zero variance and loss of determinism.

Beyond that, aggressive optimisations by the compiler can also introduce incorrectness in floating-point arithmetic~\cite{llvm-floating-point}. However, the same executable should still return the same output for identical input.

>>>>

A common misconception that is often jumped to when concerning non-determinism is the use of floating point number representation. This is an erroneous conflation with non-deterministic computational execution and we explain the reasons for this here. 
%
Floating-point representation of real numbers is limited by the fixed bit width available in the hardware, resulting in the finite precision with which numbers can be represented in a computation. 
Thus, the results of arithmetic operations must fit into the given bit width, which is achieved through rounding to the nearest representable value. This gives rise to rounding errors~\cite{FloatingPointsBook,goldberg1991every}. %

As a consequence, floating-point arithmetic is not associative. Thus, for arithmetic expressions beyond two operands, such as $x+y+z$, executing $x+(y+z)$ rather than $(x+y)+z$, can produce different results~\cite{Kapre2007}. This is because rounding is being performed at different stages in the computation, depending on the respective execution order. 
%
% Execution order changes can occur due to the use of a different compiler or as a consequence of parallelisation, e.g.\ within the runtime environment or at the hardware level. 
Execution order changes or aggressive optimisations~\cite{llvm-floating-point} can occur due to the use or settings of a compiler or as a consequence of parallelisation which can also introduce incorrectness in floating-point arithmetic, e.g.\ within the runtime environment or at the hardware level. 
%
Furthermore, floating-point calculation results may differ if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.

@@

% However, as illustrated above on the example of non-associative addition, a change of execution order can lead to sequences of floating-point operations producing different results. Thus, any uncontrolled options to modify execution order, e.g.\ at the compiler level, within the runtime environment or at the hardware level, can cause simulation results to differ between runs, resulting in non-zero variance and loss of determinism.

% Beyond that, aggressive optimisations by the compiler can also introduce incorrectness in floating-point arithmetic~\cite{llvm-floating-point}. However, the same executable should still return the same output for identical input.
